Counting evolutions: a simple foundation for thermodynamics

### 2021-02-12 Articles/2019-Entropy/Source/thermodynamics.tex
Another pass with Christine. Version sent to Mario.

To finish, consider a \textbf{deterministic processes with equilibria}, like the one in figure \ref{fig_with_equilibria}. An equilibrium for a process is a state that, once it is reached, it is never left.\footnote{This characterization of equilibrium, though amply used in thermodynamic literature, is simplistic and technically does not quite work. However, it will suffice for the purpose of this paper as a full treatment would again distract from the main points.} In a process with equilibria all evolutions will reach, at some point, an equilibrium. In the out-of-equilibrium phase, evolutions merge until they reach equilibria and can no longer merge. This means that our measure $\mu$ is maximized at equilibria. If we care to characterize only the initial and final state, deterministic evolution simply means that the final equilibrium is a function of the initial state.

@@

We already reached an important result: with only a handful of simple definitions we have found a quantity, $\mu(\lambda(t))$, that can never decrease in a deterministic process, stays the same if the process is also reversible and is maximized at equilibrium. We have done so with no assumption on the type of system, the type of forces acting on the system, the internal composition of the system or the internal dynamics of the system. It knows nothing of uncertainty, disorder, information, statistical or probability distribution, phase space volumes. It only knows about evolutions and how to count them. The result is completely general and the justification is very straight forward. It is conceptually a perfect candidate for the thermodynamic entropy, though we have one questions left: how does it combine under system composition?

Suppose we have two systems, with their respective state spaces $X$ and $Y$. The state space of the composite system is the set of all possible pairs $X \times Y$. A process defined on the composite system will come with a measure $\mu_{X \times Y}$ that enables us to count evolutions $\lambda(t) = \{\lambda_x(t), \lambda_y(t)\}$ of the composite system. In general, the evolution of one system will affect the evolution of the other system. This will not happen, however, if the systems are independent. That is, as shown in \ref{fig_independence}, for each possible evolution $\lambda_x(t)$ of the first system and $\lambda_y(t)$ of the second system, there will be a possible evolution $\{\lambda_x(t), \lambda_y(t)\}$ on the composite. In other words,
\begin{equation}
	\mu_{X \times Y} = \mu_X \mu_Y
\end{equation}
the measure factorizes. We say two systems are \textbf{independent} at a given time $t$ if the evolutions per state factorizes: $\mu_{X \times Y}(\lambda(t)) = \mu_X (\lambda_x(t)) \mu_Y (\lambda_y(t))$.

@@

This link between state entropy and independent equilibria opens a series of questions on the fundamental link between the two which go beyond the scope of this paper. However, we want at least give a partial answer to one: why is assigning a unique process entropy important? One answer is process combination. Suppose that we are studying a system and we change the circumstances surrounding it: it first evolves according to process $A$ and at $t_0$ it switches to process $B$. If $\lambda(t)$ is an evolution of the combined process $AB$, then for $t \leq t_0$ it must also be an evolution for $A$ and for $t \geq t_0$ it must also be an evolution for $B$. This means that at the junction state $x_{t_0} = \lambda(t_0)$ the count of evolution has to match: $\mu_{AB}(x_{t_0})=\mu_A(x_{t_0})=\mu_B(x_{t_0})$. If we want to mix and match from a set of possible processes, then the process entropy at the junction must be only a function of the state, independent of the process. As we'll see in section \ref{sec_mechanics}, these type of consideration are already embedded in the mathematical structure of the state spaces of classical mechanics, quantum mechanics, thermodynamics and statistical mechanics.

@@

We define a \textbf{thermodynamic process} a particular type of deterministic process with equilibria that satisfies two additional requirements:
\begin{itemize}
	\item all parts of the systems are in equilibria and are independent from each other
	\item there is a non-trivial dependence between the initial state of one part and the final state of another (i.e. there is a way to change the initial state of one part such that the final state of another is effectively changed).
\end{itemize}
The first requirement tells us that process entropy is maximized at all levels of the system. The second requirements serves to rule out the case where the systems are isolated from each other. We say a system is in \textbf{thermodynamic equilibrium} if it is part of the output of a thermodynamic process. We say two systems are in thermodynamic equilibrium with each other if they are parts of the same system in thermodynamic equilibrium.\footnote{Mathematically, the set of all systems in the same thermodynamic equilibrium is a principal ideal $X\!\! \downarrow$.}

@@

The next step is to characterize to what extent can thermodynamic processes change the energy and the entropy of each system. Intuitively, there are two ``orthogonal'' directions for a change of energy: constant entropy and maximal entropy change. To study these components, it is useful to consider the following special thermodynamic systems.

We say a thermodynamic system $R$ is a \textbf{thermal reservoir} if the internal energy $U_R$ is the only state variable. In this case, the equation of state is simply $S_R = S_R(U_R)$ and $\Delta S_R = \int \beta_R dU_R$.\footnote{The existence of unique equilibria between reservoirs forces $S_R(U_R)$ to be strictly monotonic.} We call \textbf{heat} $Q = - \Delta U_R$ the energy lost by the reservoir during a thermodynamic process.

We say $M$ is a \textbf{mechanical system} if the state entropy is a constant for all states. For a thermodynamic system that is also mechanical, we therefore have  $S_M(U_M, x^i_M) = k_M$ for all its states. A mechanical system cannot change entropy and all states can be connected through a deterministic and reversible process. We have:
\begin{equation}
\begin{aligned}
dU_M &= k_B T_M dS + X_{Mi} dx^i_M = X_{Mi} dx^i_M.
\end{aligned}
\end{equation}
The energy of the mechanical system is effectively \emph{stored} in one of the other state variables and can be later retrieved. We call \textbf{work} the energy exchanged by a mechanical system and we note $W=\Delta U_M$ as the energy gained by the system during a thermodynamic process.\footnote{Note that $\frac{\partial S_M}{\partial U_M} = 0 = \beta = \frac{1}{k_B T_M}$ which means the thermodynamic ``temperature'' of a mechanical system is infinite. While this ``temperature'' has little to do with thermometers, the fact that we can always transform all work into heat but not vice-versa is just a specific case of being able to move energy from higher to lower temperature.}

@@

The third law has many equivalent formulations. [CITE] One states that the entropy of any system has a non-negative lower bound which is reached by the lowest energy state. If the ground state is non-degenerate, such as a perfect crystal, its entropy is equal to zero. With some work, we can recover this form.

@@

First of all, the existence of unique thermodynamic equilibria forces $S(U)$ to be strictly concave (i.e. negative second derivative). This means that, at least in some region, $S(U)$ must either be a decreasing or increasing function. But since $S$ cannot be negative, $S(U)$ cannot be defined on the whole domain. Therefore $U$ must be bounded at least from above or below. Consequently, the minimum for entropy is reached either at the lowest or highest energy state. Now we show that either all thermodynamic systems must have a lowest energy state or they must have a highest energy state. Suppose, in fact, we had two systems: $X$ has no maximum energy while $Y$ has no minimum energy. Since $S_X(U_X)$ is concave, it must be a monotonically increasing function of $U_X$. This means that $\beta_X$ is always positive. Conversely, $S_X(U_X)$ must be monotonically decreasing, which means $\beta_X$ is always negative. If we put the two systems in contact, the energy of $Y$ will keep decreasing and the energy of $X$ will keep increasing and no equilibrium will ever be reached. Therefore all thermodynamic systems either have a minimum or a maximum energy. At this point the difference is a sign convention, so we can assume by convention that all thermodynamic systems have a minimum energy.

@@

We have thus recovered the laws of thermodynamics, and we have done so only by characterizing how equilibria and state variables behave under system composition. Apart from that, we have made no assumptions on the type of system, the type of forces acting on the system, the internal composition of the system or the internal dynamics of the system. No reference to classical mechanics, quantum mechanics or statistical mechanics is needed: the theory stands on its own. This fits the original spirit of thermodynamics, which was to find a set of laws that would be independent fn the ultimate dynamics of the constituents. This also explains the success of these ideas in fields outside physics. Additionally, the concepts we used are directly related to the physical explanation, and the laws are recovered immediately, not after pages and pages of calculations. The combination of these features makes this approach directly useful, for example, in an introductory  course: a coherent self-contained explanation can be given in terms of simple definitions.

@@

The second way to define entropy is through the Gibbs and von Neumann entropies, which essentially are the application of the Shannon entropy in the classical and quantum case. One way we can relate this to the process entropy is the following. Suppose we have a macroscopic equilibrium which can be understood as a dynamical equilibrium of the microscopic description. That is, the macroscopic equilibrium is a set of evolutions whose microscopic description can be characterized by a stable probability distribution $\rho(x)$. An evolution $\lambda(t)$ for the microstate, then, is an dense sequence of infinitely many microstates whose recurrence matches $\rho(x)$. Since the state entropy is the maximal process entropy, we are counting all possible evolutions which means all possible permutations. The Shannon entropy can be understood as the count of all possible permutations of an infinite sequence, of all possible evolutions.

These links are very straight forward and are fundamental in the sense that we are not applying different definitions that happen to agree: we are trivially applying the general concept, counting evolutions, in the different cases and finding the correct result. There is no mystery as to what is going on in the above two paragraphs, it's just a computation.

Note, however, that both definition relies on a somewhat implicit premise: all classical and quantum states individually have the same state entropy, they are, as we defined them, mechanical systems. In quantum mechanics, for example, the von Neumann entropy of every pure state is zero. In classical mechanics, a uniform measure is used to count states, which means all states must contribute the same entropy. Additionally, in both theories, given two states, we can always (at least mathematically) find a deterministic and reversible process (Hamiltonian evolution and unitary evolution are entropy preserving) that starts with the first and ends in the second. So the idea that mechanical systems are associated with a constant entropy, though seldom explicit, is already embedded in the established theories. In fact, all geometrical features of those spaces can be understood in terms of counting evolutions.

The geometrical structure of classical mechanics (i.e. the symplectic structure of phase space) can be understood to be exactly the structure we need to count the evolutions of a deterministic and reversible process (i.e. a Hamiltonian process). In that case state entropy and process entropy coincide, but not in general. For example, in dissipative processes (e.g. damped harmonic oscillator), this relationship breaks down. As the system loses energy, the evolutions gradually concentrate around attractors. In terms of count of states, areas shrink over time, so counting states gives us a decreasing number, which is at odds with the idea of entropy increasing when progressing towards an equilibrium. On the other hand, each region as it evolves will correspond to the same number of evolutions, which will now be spread over fewer states: the process entropy (i.e. the logarithm of the evolutions per state) increases as evolutions become more concentrated. In the same way that thermodynamic entropy is the process entropy under isoentropic evolution (i.e. at equilibrium), the state entropy in classical mechanics is the process entropy under isoentropic evolution (i.e. deterministic and reversible evolution). Therefore, in the same way thermodynamic entropy does not apply outside of equilibrium, one has to be careful to apply the symplectic structure outside of Hamiltonian evolution.

The mathematical structure of quantum mechanics also can be understood in terms of evolution counts. The inner product of quantum mechanics gives us the conditional probabilities of a ``black box'' process (i.e. the projection/measurements) that, given an initial state, gives us a final one. As we saw, conditional probabilities are ratios of evolution counts: what fractions of the evolutions that start in one state end in another. In other words, the inner product tells us how the evolutions coming in and coming out of these processes are to be ``joined'' with the preceding and following ones.  Unitary evolution is exactly when all evolution of one state go to one and only one state.  Moreover, note that these ``black box'' processes are projections, meaning they give the same result if applied more than once. Eigenstates of projections, then, are equilibria of the process the projection represents. Or, in other words, equilibria are symmetries of the process.

The above insights show that the concepts we are using do not just provide a nicer retelling of thermodynamics, but they relate to something more fundamental that underlies all physical theories. We believe that this is not a coincidence, but stems from an obvious requirement that we must put on physical theories. Most, if not all, experiments are described as a sequence of different steps (e.g. one or more preparation, manipulation and measurement) stitched together. Therefore a theory must be able to tell us how to ``stitch together'' the evolutions at every junction, how to keep the evolution count correctly. This bookkeeping is a requirement and therefore it makes sense it plays such a major role in defining state spaces.

@@

In this paper we gave a conceptual overview for a new approach to recover thermodynamics based on the simple idea of counting evolutions, the possible ways a system can evolve under a process. We saw that the logarithm of evolution count under a deterministic process has all the properties we associate with entropy: it cannot decrease over time, it is constant in reversible processes and it is maximized at equilibrium. The state entropy represents the maximized process entropy at equilibrium and for the composite of independent systems. Thermodynamic processes and equilibria describe the case where the equilibrium is achieved at all levels: the whole and any of the parts. This gives a simple way to rederive thermodynamics in a straight-forward and conceptually insightful way.

We have also seen that two main ways to define entropy in statistical mechanics can be understood as the process entropy for the ensembles, and that the geometrical properties of classical and quantum mechanics can be understood in terms of evolutions counts. Keeping track of evolution counts is fundamental in a physical theory for process composition, as we need to make sure that the evolutions are ``stitched together'' correctly when processes are joined. Our hope and long term goal is to develop a formal theory based on these ideas that identifies a minimum set of requirements for the definition of a state space. This basic definition can then be further refined with different additional assumptions, each set recovering the different physical theories. We believe this would help us better understand the right questions to ask in within the foundations of physics.

>>>>

To finish, consider a \textbf{deterministic processes with equilibria}, like the one in figure \ref{fig_with_equilibria}. An equilibrium for a process is a state that, once it is reached, it is never left.\footnote{This characterization of equilibrium, though amply used in thermodynamic literature, is simplistic and technically does not quite work, e.g. is not relativistically invariant. However, it will suffice for the purpose of this paper as a full treatment would again distract from the main points.} In a process with equilibria all evolutions will reach, at some point, an equilibrium. In the out-of-equilibrium phase, evolutions merge until they reach equilibria and can no longer merge. This means that our measure $\mu$ is maximized at equilibria. If we care to characterize only the initial and final state, deterministic evolution simply means that the final equilibrium is a function of the initial state.

@@

We already reached an important result: with only a handful of simple definitions we have found a quantity, $\mu(\lambda(t))$, that can never decrease in a deterministic process, stays the same if the process is also reversible and is maximized at equilibrium. We have done so with no assumption on the type of system, the type of forces acting on the system, the internal composition of the system or the internal dynamics of the system. It knows nothing of uncertainty, disorder, information, statistical or probability distribution, phase space volumes. It only knows about evolutions and how to count them. The result is completely general and the justification is very straightforward. It is conceptually a perfect candidate for the thermodynamic entropy, though we have one question left: how does it combine under system composition?

Suppose we have two systems, with their respective state spaces $X$ and $Y$. The state space of the composite system is the set of all possible pairs $X \times Y$. A process defined on the composite system will come with a measure $\mu_{X \times Y}$ that enables us to count evolutions $\lambda(t) = \{\lambda_x(t), \lambda_y(t)\}$ of the composite system. In general, the evolution of one system will affect the evolution of the other system. This will not happen, however, if the systems are independent. That is, as shown in figure \ref{fig_independence}, for each possible evolution $\lambda_x(t)$ of the first system and $\lambda_y(t)$ of the second system, there will be a possible evolution $\{\lambda_x(t), \lambda_y(t)\}$ on the composite. In other words,
\begin{equation}
	\mu_{X \times Y} = \mu_X \mu_Y
\end{equation}
the measure factorizes. We say two systems are \textbf{independent} at a given time $t$ if the evolutions per state factorizes: $\mu_{X \times Y}(\lambda(t)) = \mu_X (\lambda_x(t)) \mu_Y (\lambda_y(t))$.

@@

This link between state entropy and independent equilibria opens a series of questions on the fundamental link between the two which go beyond the scope of this paper. However, we want at least to give a partial answer to one: why is assigning a unique process entropy important? One answer is process combination. Suppose that we are studying a system and we change the circumstances surrounding it: it first evolves according to process $A$ and at $t_0$ it switches to process $B$. If $\lambda(t)$ is an evolution of the combined process $AB$, then for $t \leq t_0$ it must also be an evolution for $A$ and for $t \geq t_0$ it must also be an evolution for $B$. This means that at the junction state $x_{t_0} = \lambda(t_0)$ the count of evolutions has to match: $\mu_{AB}(x_{t_0})=\mu_A(x_{t_0})=\mu_B(x_{t_0})$. If we want to mix and match from a set of possible processes, then the process entropy at the junction must be only a function of the state, independent of the process. As we'll see in section \ref{sec_mechanics}, these types of consideration are already embedded in the mathematical structure of the state spaces of classical mechanics, quantum mechanics, thermodynamics and statistical mechanics.

@@

We define a \textbf{thermodynamic process} a particular type of deterministic process with equilibria that satisfies two additional requirements:
\begin{itemize}
	\item in the final equilibrium, all parts of the system are independent from each other and are themselves in equilibrium
	\item there is a non-trivial dependence between the initial state of one part and the final state of another (i.e. there is a way to change the initial state of one part such that the final state of another is effectively changed).
\end{itemize}
The first requirement tells us that process entropy is maximized at all levels of the system. The second requirement serves to rule out the case where the systems are isolated from each other. We say a system is in \textbf{thermodynamic equilibrium} if it is part of the output of a thermodynamic process. We say two systems are in thermodynamic equilibrium with each other if they are parts of the same system in thermodynamic equilibrium.\footnote{Mathematically, the set of all systems in the same thermodynamic equilibrium is a principal ideal $X\!\! \downarrow$.}

@@

For a thermodynamic system, then, the state entropy can be expressed as $S(U, x^i)$ a function of the extensive quantities.\footnote{The existence of unique equilibria between thermodynamic systems forces $S$ to be strictly monotonic in $U$.} This is simply the equation of state in entropic form, which is the starting point of Gibbs' formulation of thermodynamics. %[XXX: CITE]
We can therefore define the relationships
\begin{align}
\beta &= \frac{\partial S}{\partial U} = \frac{1}{k_B T} \;\;\;\;\;\; - \beta X_i = \frac{\partial S}{\partial x^i}
\end{align}
where $k_B$ is Boltzmann's constant. We find:
\begin{align}
dS &= \frac{\partial S}{\partial U} dU + \frac{\partial S}{\partial x^i} dx^i \\
&= \beta dU - \beta X_i dx^i \\
dU &= k_B T dS + X_i dx^i
\end{align}
which are the usual thermodynamic relationships, except that, for us, the state entropy $S$ is dimensionless.

The next step is to characterize to what extent thermodynamic processes can change the energy and the entropy of each system. Intuitively, there are two ``orthogonal'' directions for a change of energy: constant entropy and maximal entropy change. To study these components, it is useful to consider the following special thermodynamic systems.

We say a thermodynamic system $R$ is a \textbf{thermal reservoir} if the internal energy $U_R$ is the only state variable. In this case, the equation of state is simply $S_R = S_R(U_R)$ and $\Delta S_R = \int \beta_R dU_R$. We call \textbf{heat} $Q = - \Delta U_R$ the energy lost by the reservoir during a thermodynamic process.

We say $M$ is a \textbf{mechanical system} if the state entropy is a constant for all states. For a thermodynamic system that is also mechanical, we therefore have  $S_M(U_M, x^i_M) = k_M$ for all its states for some constant $k_M$. A mechanical system cannot change entropy and all states can be connected through a deterministic and reversible process. We have:
\begin{equation}
\begin{aligned}
dU_M &= k_B T_M dS_M + X_{Mi} dx^i_M = X_{Mi} dx^i_M.
\end{aligned}
\end{equation}
The energy of the mechanical system is effectively \emph{stored} in one of the other state variables and can be later retrieved. We call \textbf{work} the energy exchanged by a mechanical system and we note $W=\Delta U_M$ as the energy gained by the system during a thermodynamic process.\footnote{Note that $\frac{\partial S_M}{\partial U_M} = 0 = \beta_M = \frac{1}{k_B T_M}$ which means the thermodynamic ``temperature'' of a mechanical system is infinite. While this ``temperature'' has little to do with thermometers, the fact that we can always transform all work into heat but not vice-versa is just a specific case of being able to move energy from higher to lower temperature.}

@@

The third law has many equivalent formulations. %[CITE]
One states that the entropy of any system has a non-negative lower bound which is reached by the lowest energy state. If the ground state is non-degenerate, such as a perfect crystal, its entropy is equal to zero. With some work, we can recover this form.

@@

First of all, the existence of unique thermodynamic equilibria forces $S(U)$ to be strictly concave (i.e. negative second derivative). This means that, at least in some region, $S(U)$ must either be a decreasing or increasing function. But since $S$ cannot be negative, $S(U)$ cannot be defined on the whole domain. Therefore $U$ must be bounded at least from above or below. Consequently, the minimum for entropy is reached either at the lowest or highest energy state. Now we show that either all thermodynamic systems must have a lowest energy state or they must have a highest energy state. Suppose, in fact, we had two systems: $X$ has no maximum energy while $Y$ has no minimum energy. Since $S_X(U_X)$ is concave, it must be a monotonically increasing function of $U_X$. This means that $\beta_X$ is always positive. Conversely, $S_Y(U_Y)$ must be monotonically decreasing, which means $\beta_Y$ is always negative. If we put the two systems in contact, the energy of $Y$ will keep decreasing and the energy of $X$ will keep increasing and no equilibrium will ever be reached. Therefore all thermodynamic systems either have a minimum or a maximum energy. At this point the difference is a sign convention, so we can assume by convention that all thermodynamic systems have a minimum energy.

@@

We have thus recovered the laws of thermodynamics, and we have done so only by characterizing how equilibria and state variables behave under system composition. Apart from that, we have made no assumptions on the type of system, the type of forces acting on the system, the internal composition of the system or the internal dynamics of the system. No reference to classical mechanics, quantum mechanics or statistical mechanics is needed: the theory stands on its own. This fits the original spirit of thermodynamics, which was to find a set of laws that would be independent of the ultimate dynamics of the constituents. This also explains the success of these ideas in fields outside physics. Additionally, the concepts we used are directly related to the physical explanation, and the laws are recovered immediately, not after pages and pages of calculations. The combination of these features makes this approach directly useful, for example, in an introductory  course: a coherent self-contained explanation can be given in terms of simple definitions.

@@

The second way to define entropy is through the Gibbs and von Neumann entropies, which essentially are the application of the Shannon entropy in the classical and quantum case. One way we can relate this to the process entropy is the following. Suppose we have a macroscopic equilibrium which can be understood as a dynamical equilibrium of the microscopic description. That is, the macroscopic equilibrium is a set of evolutions whose microscopic description can be characterized by a stable probability distribution $\rho(x)$. An evolution $\lambda(t)$ for the microstate, then, is a dense sequence of infinitely many microstates whose recurrence matches $\rho(x)$. Since the state entropy is the maximal process entropy, we are counting all possible evolutions which means all possible permutations. The Shannon entropy can be understood as the count of all possible permutations of an infinite sequence, of all possible evolutions.

These links are very straightforward and are fundamental in the sense that we are not applying different definitions that happen to agree: we are trivially applying the general concept, counting evolutions, in the different cases and finding the correct result. There is no mystery as to what is going on in the above two paragraphs, it's just a computation.

Note, however, that both derivations rely on a somewhat implicit premise: all classical and quantum states individually have the same state entropy; they are mechanical systems as we defined them. In quantum mechanics, for example, the von Neumann entropy of every pure state is zero. In classical mechanics, a uniform measure is used to count states, which means all states must contribute the same entropy. Additionally, in both theories, given two states, we can always (at least mathematically) find a deterministic and reversible process (Hamiltonian evolution and unitary evolution are entropy preserving) that starts with the first and ends in the second. So the idea that mechanical systems are associated with a constant entropy, though seldom explicit, is already embedded in the established theories. In fact, all geometrical features of those spaces can be understood in terms of counting evolutions.

The geometrical structure of classical mechanics (i.e. the symplectic structure of phase space) can be understood to be exactly the structure we need to count the evolutions of a deterministic and reversible process (i.e. a Hamiltonian process). In that case state entropy and process entropy coincide, but not in general. For example, in dissipative processes (e.g. a damped harmonic oscillator), this relationship breaks down. As the system loses energy, the evolutions gradually concentrate around attractors. In terms of count of states, areas shrink over time, so counting states gives us a decreasing number, which is at odds with the idea of entropy increasing when progressing towards an equilibrium. On the other hand, each region as it evolves will correspond to the same number of evolutions, which will now be spread over fewer states: the process entropy (i.e. the logarithm of the evolutions per state) increases as evolutions become more concentrated. In the same way that thermodynamic entropy is the process entropy under isentropic evolution (i.e. at equilibrium), the state entropy in classical mechanics is the process entropy under isentropic evolution (i.e. deterministic and reversible evolution). Therefore, in the same way thermodynamic entropy does not apply outside of equilibrium, one has to be careful when trying to apply the symplectic structure outside of Hamiltonian evolution.

The mathematical structure of quantum mechanics can also be understood in terms of evolution counts. The inner product of quantum mechanics gives us the conditional probabilities of a ``black box'' process (i.e. the projection/measurements) that, given an initial state, gives us a final one. As we saw, conditional probabilities are ratios of evolution counts: what fractions of the evolutions that start in one state end in another. In other words, the inner product tells us how the evolutions coming in and going out of these processes are to be ``joined'' with the preceding and following ones.  Unitary evolution is exactly when all evolutions of one state go to one and only one state.  Moreover, note that these ``black box'' processes are projections, meaning they give the same result if applied more than once. Eigenstates of projections, then, are equilibria of the process the projection represents. Or, in other words, equilibria are symmetries of the process.

The above insights show that the concepts we are using do not just provide a nicer retelling of thermodynamics, but they relate to something more fundamental that underlies all physical theories. We believe that this is not a coincidence, but stems from an obvious requirement that we must put on physical theories. Most, if not all, experiments are described as a sequence of different steps (e.g. one or more preparations, manipulations and measurements) stitched together. Therefore a theory must be able to tell us how to ``stitch together'' the evolutions at every junction, how to keep the evolution count correctly. This bookkeeping is a requirement and therefore it makes sense it plays such a major role in defining state spaces.

@@

In this paper we gave a conceptual overview for a new approach to recover thermodynamics based on the simple idea of counting evolutions, the possible ways a system can evolve under a process. We saw that the logarithm of evolution count under a deterministic process has all the properties we associate with entropy: it cannot decrease over time, it is constant in reversible processes and it is maximized at equilibrium. The state entropy represents the maximized process entropy at equilibrium and for the composite of independent systems. Thermodynamic processes and equilibria describe the case where the equilibrium is achieved at all levels: the whole and any of the parts. This gives a simple way to rederive thermodynamics in a straightforward and conceptually insightful way.

We have also seen that two main ways to define entropy in statistical mechanics can be understood as the process entropy for the ensembles, and that the geometrical properties of classical and quantum mechanics can be understood in terms of evolution counts. Keeping track of evolution counts is fundamental in a physical theory for process composition, as we need to make sure that the evolutions are ``stitched together'' correctly when processes are joined. Our hope and long term goal is to develop a formal theory based on these ideas that identifies a minimum set of requirements for the definition of a state space. This basic definition can then be further refined with different additional assumptions, each set recovering the different physical theories. We believe this would help us better understand the right questions to ask within the foundations of physics.
